---
title: "Traffic Fatalities"
author: "Blake, Charlie, Eshna, Noah"
date: "February 12, 2021"
output: bookdown::html_document2
df_print: paged
number_sections: yes
fig_caption: yes

---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, message=FALSE, results='hide'}
# clear environment
rm(list=ls())

# Clear all plotsfee
try(dev.off(dev.list()["RStudioGD"]),silent=TRUE)
try(dev.off(),silent=TRUE)

# load packages
library(tidyverse)
library(haven)
library(car)
library(gtsummary)
```

# Abstract 

# Introduction



# Background 

# Descriptive Analysis 

# Statistical Approach

There are three main linear regression methods for analyzing longitudinal data: 1) fixed effects (FE), 2) random effects (RE), and 3) pooled ordinary least squares (OLS) (Dougherty, 2007). The third may only be used if the control variables are exhaustively comprehesive so that they capture all of the relevant variation between the entities (states), which is certainly not the case here (Dougherty, 2007). The second (RE) may be used when the observations in the data consist of a random sample from a larger population; since the entities in this case are not randomly selected states this rules out using RE regression (Dougherty, 2007). The first (FE) is highly attractive in that it allows for the elimination of unobserved, time-invariant heterogeneity between entities that might induce bias in the regression (Dougherty, 2007). There are three versions of FE linear regression, they are 1) within-groups, 2) first differences, and 3) least squares dummy variable (LSDV) (Dougherty, 2007). This analysis uses the third option, which is technically equivalent to the first, due to ease of implementation and interpretation of results in R. The standard specification of a linear longitudinal model is given by Dougherty (2007):

$Y_{it} = \beta_1 + \sum^{k}_{j=2}\beta_jX_{jit}+ \sum^{s}_{p=1}\gamma_pZ_{pi} + \delta t + \epsilon_{it}$

where "Y is the dependent variable, $X_{j}$ are observed explanatory variables, and the $Z_{p}$ are unobserved explanatory variables" (Dougherty, 2007). The index $i$ indicates the entity, $t$ indicates the time period, $j$ and $p$ indicate different explanatory variables, $\delta t$ is a trend term to allow the intercept to shift over time and $\epsilon_{it}$ is the disturbance term (Dougherty, 2007). In this analysis a full set of  year dummy variables is included so the trend term is unecessary.

To deal with the $Z_{p}$, which are unobservable, time-invariant characteristics of each state that induce bias, its effects can be lumped into a component of the error term, as shown below.

$Y_{it} = \beta_1 + \sum^{k}_{j=2}\beta_jX_{jit}+ \alpha_i + \delta t + \epsilon_{it}$

With

$\alpha_i = \sum_{p=1}^{s}\gamma_p Z_p$

Unfortunately the presence of $\alpha_i$ will in general result in inefficient OLS estimates, but the three methods listed in the beginning of this section overcome this challenge. The chosen method, LSDV OLS, is shown below for the implementation in this analysis.

$Y_{it} = \sum^{k}_{j=2}\beta_jX_{jit}+ \sum_{i=1}^{n}\alpha_i A_i + \epsilon_{it}$

Where the unobserved effect $\alpha_i$ is now distributed as the coefficient of the individual-specific dummy variable. As Dougherty (2007) puts it, "the $\alpha_i A_i$ term represent[s] [the] fixed effect on the dependent variable $Y_i$ for individual $i$. This model can now be estimated using OLS (Dougherty, 2007). Assuming there are no omitted and relevant time-varying explanatory variables, this model will give unbiased results.

Problems with the LSDV regression approach are identical to those of the within-groups method and include 1) inability to estimate the effects of time-invariant explanatory variables, 2) the disturbance term can be small compared to the varation in the differenced values (in within-groups the model is differenced to eliminate the time-invariant explanatory variables), 3) one degree of freedom is lost for every entity (Dougherty, 2007).

Our end goal is to find a policy suggestion that can decrease traffic fatalities. However, the Fatalities dataset primarily contains variables related to alcohol-related policies. Among the various measures of vehicle fatalities in our data, we choose alcohol-involved vehicle fatalities per 100,000 people in the population as the response variable as it is the metric of fatalities that is most closely-related to the set of explanatory variables we have to choose from. We use alcohol fatalities per 100,000 instead of per capita for ease of interpretation of the coefficients. Furthermore, alcohol-involved vehicle fatalities per 100,000 has a strong positive correlation with the total vehicle fatalities per 100,000. Therefore, we do not expect the choice of response variable to substantively affect our conclusions.

We consider three different groups of explanatory variables: state fixed effect, year fixed effects, and various policy-related variables. We choose to include state fixed effects in our model to account for the variation in a multitude of state-specific factors that are not included in our data such as climate, geography, and culture. For example, high levels of alcohol consumption in New York City, where there is easy access to public transportation, may present less of a threat to road safety than in rural Wisconsin where it may be more likely to result in drunk driving. Next, we include dummy variables for each year to account for any time trends in the data. Finally, we included the various policy-related variables in our data: change in beer tax (`beertax_change`), minimum legal drinking age (`drinkage`), a dummy variable for whether a preliminary breath test law exists (`breath`), a dummy variable for whether there is a mandatory jail sentence law (`jail`), and a dummy variable for whether there is a mandatory community service law (`service`) to determine if there is any evidence that these policies are associated with lower fatalities. We choose to include the change in beer tax in each state instead of the actual beer tax to reduce the multicollinearity between the beer tax variable and state dummies.


# Results

**Regression Results**
```{r, echo = FALSE}
data(Fatalities, package = "AER")

beertax_1982 <- Fatalities[Fatalities$year == 1982, c('state', 'beertax')]
colnames(beertax_1982) <- c('state', 'beertax_1982')

Fatalities <- merge(Fatalities, beertax_1982, by = 'state')
Fatalities$beertax_change <- Fatalities$beertax - Fatalities$beertax_1982

fit1 <- lm(afatal/pop*100000 ~ + state + year + beertax_change + drinkage + breath + jail 
          + service  - 1, 
          data = Fatalities)

summary(fit1)$coefficients[55:59,]
  # not reporting states
```

From the table of coefficients above, we see that among the only policy-related variable with a negative statistically significant estimated coefficient at the 5 percent significance level is the change in beer tax. Thus, we conclude that the policy that is most strongly associated with a decrease in alcohol-related fatalities is an increase in beer tax. We can interpret the coefficient on the `beertax_change` variable as holding all else equal, a one unit increase in the tax on a case of beer on average is associated with a decrease of 2.92 in alcohol-related fatalities per 100,000. 

The estimated coefficient on the dummy variable on the mandatory jail sentence dummy is also individually statistically significant at the 5 percent level. However, the coefficient is positive, which goes against our intuition that increased penalties would be associated with lower fatalities, so we do not make any policy suggestion related to jail sentencing. 

The estimated coefficients on the minimum legal drinking age and the preliminary breath test law dummy variable are very small in magnitude and not individually statistically significant at the 5 percent level. The mandatory community service dummy variable (`service`) is larger and negative, which catches our attention, but the p-value is also relatively high, so we do not find a convincing association between mandatory community service and decreased alcohol-involved vehicle fatalities.

The estimated coefficients on the state and year fixed effects are not shown as there are many and we do not draw any important conclusions related to our question of interest from these coefficients.

As an additional sensitivity, we try running a model with the same set of explanatory variables but using the vehicle fatalities per 100,000 as the response variable instead of alcohol-related fatalities per 100,000.


# Sensitivity Analysis

As an additional sensitivity, I try running model with the same set of explanatory variables but using the vehicle fatalities per 100,000 as the response variable instead of alcohol-related fatalities per 100,000.


```{r, echo = FALSE}
fit2 <- lm(fatal/pop*100000 ~ + state + year + beertax_change + drinkage + breath + jail 
          + service - 1, 
          data = Fatalities)

summary(fit2)$coefficients[55:59,]
  # not reporting states
```

In the table above we see that the coefficients have changed. However, the policy for which there is strongest evidence of association with decreased fatalities is still an increase in beer tax, which indicates that our conclusions are not sensitive to our choice of alcohol-involved vehicle fatalities per 100,000 instead of overall vehicle fatalities per 100,000 as the response variable.

The Residuals vs Fitted plots visually shows that the residuals are mostly around the 0 line. This suggests that the assumption of using a linear relationship is reasonable even though there are outliers present.

```{r,echo=FALSE,include=TRUE}
plot(fit1,which=1)
```

In the QQ Plot, we can see that points are mostly in a line at the middle of the graph and curves off with extreme points.
This means that the data set contains more extreme values than expected when compared to a Normal distribution.


```{r,echo=FALSE,include=TRUE}
plot(fit1,which=2)
```

It is evident that there are outliers in this plot.

```{r,echo=FALSE,include=TRUE}
plot(fit1,which=3)
```




```{r,echo=FALSE,include=TRUE}
plot(fit1,which=5)
```

# Discussion

# Conclusions

# Acknowledgements {-}


# References {-}

Dougherty, C. (2007). Introduction to Econometrics (3rd ed., p. 480). Oxford: Oxford University Press.

# Appendix A

# Appendix B

# Session info {-}

```{r}
sessionInfo()
```